### 2015-12-13 ada-ftrl-proofs.tex
Fix typos, grammar, and bibliography.

\begin{proof}[Proof of Lemma~\ref{lemma:recurrence-solution}]
We prove the lemma by induction on $T$.  The base case $T=0$ is trivial. For $T
\ge 1$, we have
$$
\Delta_T
\le \Delta_{T-1} + \min \left\{a_T, \ \frac{a_T^2}{\Delta_{T-1}} \right\}
\le \sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
$$
where the first inequality follows from Lemma~\ref{lemma:gap-recurrence}, and
the second inequality from the induction hypothesis and the fact that $f(x) = x
+ \min\{a_T, a_T^2/x\}$ is an increasing function of $x$.  It remains to prove
that
$$
\sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
\le  \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
Dividing through by $a_T$ and making substitution $z=\frac{\sqrt{\sum_{t=1}^{T-1} a_t^2}}{a_T}$, leads to
$$
z\sqrt{3} + \min\left\{1,\frac{1}{z\sqrt{3}}\right\} \le \sqrt{3 + 3z^2}
$$
which can be easily checked by considering separately the cases $z \in
[0,\frac{1}{\sqrt{3}})$ and $z \in [\frac{1}{\sqrt{3}}, \infty)$.
\end{proof}

>>>>

\begin{proof}[Proof of Lemma~\ref{lemma:recurrence-solution}]
We prove the lemma by induction on $T$.  The base case $T=0$ is trivial. For $T
\ge 1$, we have
$$
\Delta_T
\le \Delta_{T-1} + \min \left\{a_T, \ \frac{a_T^2}{\Delta_{T-1}} \right\}
\le \sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
$$
where the first inequality follows from Lemma~\ref{lemma:gap-recurrence}, and
the second inequality from the induction hypothesis and the fact that the
function $f(x) = x + \min\{a_T, a_T^2/x\}$ is increasing on $[0,\infty)$.  It
remains to prove that
$$
\sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
\le  \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
If $a_T = 0$, the inequality clearly holds. Dividing through by $a_T$ and
making substitution $z=\frac{\sqrt{\sum_{t=1}^{T-1} a_t^2}}{a_T}$, leads to
$$
z\sqrt{3} + \min\left\{1,\frac{1}{z\sqrt{3}}\right\} \le \sqrt{3 + 3z^2}
$$
which can be easily checked by considering separately the cases $z \in
[0,\frac{1}{\sqrt{3}})$ and $z \in [\frac{1}{\sqrt{3}}, \infty)$.
\end{proof}

### 2015-12-13 conclusions.tex
Fix typos, grammar, and bibliography.

Similar, but weaker result holds for a scale-free algorithm based on
\textsc{Mirror Descent}. However, we have also shown this algorithm is strictly
weaker than algorithms based on \textsc{Follow The Regularizer Leader}. Namely,
we gave examples of regularizers for which the associated Bregman divergence is
unbounded and the scale-free version of \textsc{Mirror Descent} has $\Omega(T)$
regret or worse.

>>>>

Similar, but weaker result holds for a scale-free algorithm based on
\textsc{Mirror Descent}. However, we have also shown this algorithm is strictly
weaker than algorithms based on \textsc{Follow The Regularizer Leader}. Namely,
we gave examples of regularizers for which the scale-free version of
\textsc{Mirror Descent} has $\Omega(T)$ regret or worse.

### 2015-12-13 introduction.tex
Fix typos, grammar, and bibliography.

OLO is a fundamental problem in machine
learning~\cite{Cesa-Bianchi-Lugosi-2006, Rakhlin-Sridharan-2009,
Shalev-Shwartz-2011}.  Many learning problems can be directly phrased as OLO,
e.g., learning with expert advice~\cite{Littlestone-Warmuth-1994, Vovk-1998,
Freund-Schapire-1997, Cesa-Bianchi-Haussler-Helmbold-Schapire-Warmuth-1997},
online combinatorial optimization~\cite{Kalai-Vempala-2005,
Helmbold-Warmuth-2009, Koolen-Warmuth-Kivinen-2010}. Other problems can be
reduced to OLO, e.g. online convex
optimization~\cite{Abernethy-Bartlett-Rakhlin-Tewari-2008},
\cite[Chapter~2]{Shalev-Shwartz-2011}, online classification
~\cite{Rosenblatt-1958,Freund-Schapire-1999} and
regression~\cite{Kivinen-Warmuth-1997},
~\cite[Chapters~11~and~12]{Cesa-Bianchi-Lugosi-2006}, multi-armed
problems~\cite[Chapter~6]{Cesa-Bianchi-Lugosi-2006},
\cite{Abernethy-Hazan-Rakhlin-2008, Bubeck-Cesa-Bianchi-2012}, and batch
and stochastic optimization of convex functions~\cite{Nemirovski-Yudin-1983,
Bubeck-2015}.  Hence, a result in OLO immediately implies other results in all
these domains.

@@

For this reason, most of the existing OLO algorithms receive as an input---or
explicitly assume---an upper bound $B$ on the norm of the loss vectors.  The
input $B$ is often disguised as the learning rate, the regularization
parameter, or the parameter of strong convexity of the regularizer. Examples of
such algorithms include the \textsc{Hedge} algorithm or online projected
gradient descent with fixed learning rate.  However, these algorithms have two
obvious drawbacks.

@@

% \begin{definition}[Scale-Free Algorithm]
% \label{definition:scale-free-algorithm}
% An algorithm for OLO is \emph{scale-free} if for any sequence of loss vectors
% $\{\ell_t\}_{t=1}^\infty$ and any real number $c > 0$, on the sequences of
% $\{\ell_t\}_{t=1}^\infty$ and $\{c \cdot \ell_t\}_{t=1}^\infty$, the algorithm
% chooses the same sequence of points $\{w_t\}_{t=1}^\infty$.
% \end{definition}

@@

Both \textsc{FTRL} and \textsc{MD} are parametrized by a function called a
\emph{regularizer}. The function is assumed to be strongly convex with respect
to a given norm $\|\cdot\|$. Based on different regularizers different
algorithms with different properties can be instantiated. A summary of
algorithms for OLO is presented in Table~\ref{table:results}.  All of them,
except for \textsc{Hedge}, are instances of \textsc{FTRL} or \textsc{MD}.

Scale-free versions of \textsc{MD} include \textsc{AdaGrad
MD}~\cite{Duchi-Hazan-Singer-2011} and~\textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}. However, these algorithms have non-trivial
regret bounds only when the Bregman divergence associated with the regularizer
is bounded, which happens only when the decision is bounded as well. As we show
in Section~\ref{subsection:mirror-descent-lower-bound}, this is unavoidable,
since \textsc{AdaGrad MD} and similar algorithms based on \textsc{MD} can have
$\Omega(T)$ regret if Bregman divergence is not bounded.

@@

For unbounded decision sets no scale-free algorithm with a non-trivial regret
bound was known. Unbounded decision sets are practically important, since
learning large-scale linear models (e.g. logistic regression) can be reduced to
OLO with $\R^d$ as the decision set. One of our contributions is an
\textsc{FTRL}-based scale-free algorithm with a non-trivial regret bound for
any regularizer and any decision set, bounded or unbounded.

% While the scale-free property has been looked at in the expert setting, in
% the general OLO setting this issue has been largely ignored.  In particular,
% the \textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014}
% algorithm, for prediction with expert advice, is specifically designed to be
% scale-free.  A notable exception in the OLO literature is the discussion of
% the ``off-by-one'' issue in~\cite{McMahan-2014}, where it is explained that
% even the popular \textsc{AdaGrad} algorithm~\cite{Duchi-Hazan-Singer-2011} is
% not completely adaptive; see also our discussion in
% Section~\ref{section:solo-ftrl}. In particular, existing scale-free
% algorithms cover only some norms/regularizers and \emph{only} bounded
% decision sets. The case of \textbf{unbounded decision sets}, practically the
% most interesting one for machine learning applications, remains completely
% unsolved.

@@

We prove that the regret of \textsc{AdaFTRL} and \textsc{SOLO FTRL} after $T$
rounds is bounded by $O \left(\sqrt{\sup_{v \in K} f(v)
\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $f$ is a non-negative regularizer that is
$1$-strongly convex with respect to a norm $\|\cdot\|$ and $\|\cdot\|_*$ is its
dual norm. For \textsc{Scale-Free MD}, we prove $O \left(\sqrt{\sup_{u,v \in K}
B_f(u,v) \sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $B_f$ is the Bregman
divergence associated with a $1$-strongly convex regularizer $f$. In
Section~\ref{section:lower-bound}, we show that the $\sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$ term in the bounds term is necessary by proving
$\frac{D}{\sqrt{8}} \sqrt{\sum_{t=1}^T\|\ell_t\|_*^2}$ lower bound on the
regret of any algorithm for OLO for any decision set with diameter $D$ with
respect to the primal norm $\|\cdot\|$.

For \textsc{SOLO FTRL}, we prove that regret against a competitor $u \in K$ is
at most $O \left(f(u) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} + \max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} \right)$.  Same as before, $f$ is a non-negative
$1$-strongly convex regularizer. The decision can be bounded or unbounded. This
is the \textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial upper bound.

@@

Our proofs of the regret upper bounds rely on homogeneous inequalities
(Lemmas~\ref{lemma:recurrence-solution}, \ref{lemma:useful},
\ref{lemma:sum-of-square-roots-inverses}).  Previously, many proofs of regret
bounds relied on ad-hoc non-homogeneous inequalities, or the existing
algorithms used quantities (e.g.  $\sum_{t=1}^T \|\ell_t\|_*^2$) that are not
available until all loss vectors are processed. Two of the inequalities
(Lemmas~\ref{lemma:recurrence-solution} and \ref{lemma:useful}) are new and
interesting in their own right.

>>>>

OLO is a fundamental problem in machine
learning~\cite{Cesa-Bianchi-Lugosi-2006, Rakhlin-Sridharan-2009,
Shalev-Shwartz-2011}.  Many learning problems can be directly phrased as OLO,
e.g., learning with expert advice~\cite{Littlestone-Warmuth-1994, Vovk-1998,
Freund-Schapire-1997, Cesa-Bianchi-Haussler-Helmbold-Schapire-Warmuth-1997},
online combinatorial optimization~\cite{Kalai-Vempala-2005,
Helmbold-Warmuth-2009, Koolen-Warmuth-Kivinen-2010}. Other problems can be
reduced to OLO, e.g. online convex
optimization~\cite{Abernethy-Bartlett-Rakhlin-Tewari-2008},
\cite[Chapter~2]{Shalev-Shwartz-2011}, online classification
~\cite{Rosenblatt-1958, Freund-Schapire-1999} and
regression~\cite{Kivinen-Warmuth-1997},
~\cite[Chapters~11~and~12]{Cesa-Bianchi-Lugosi-2006}, multi-armed
problems~\cite[Chapter~6]{Cesa-Bianchi-Lugosi-2006},
\cite{Abernethy-Hazan-Rakhlin-2008, Bubeck-Cesa-Bianchi-2012}, and batch and
stochastic optimization of convex functions~\cite{Nemirovski-Yudin-1983,
Bubeck-2015}.  Hence, a result in OLO immediately implies other results in all
these domains.

@@

For this reason, most of the existing OLO algorithms receive as an input---or
explicitly assume---an upper bound $B$ on the norm of the loss vectors.  The
input $B$ is often disguised as the learning rate, the regularization
parameter, or the parameter of strong convexity of the regularizer. Examples of
such algorithms include the \textsc{Hedge}~\cite{Freund-Schapire-1997} or
online projected gradient descent with fixed learning rate.  However, these
algorithms have two obvious drawbacks.

@@

@@

Both \textsc{FTRL} and \textsc{MD} are parametrized by a function called a
\emph{regularizer}. The function is assumed to be strongly convex with respect
to a given norm. Based on different regularizers different algorithms with
different properties can be instantiated. A summary of algorithms for OLO is
presented in Table~\ref{table:results}.  All of them, except for
\textsc{Hedge}, are instances of \textsc{FTRL} or \textsc{MD}.

Scale-free versions of \textsc{MD} include \textsc{AdaGrad
MD}~\cite{Duchi-Hazan-Singer-2011} and~\textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}. However, these algorithms have non-trivial
regret bounds only when the Bregman divergence associated with the regularizer
is bounded, which happens only when the decision set is bounded as well. As we
show in Section~\ref{subsection:mirror-descent-lower-bound}, this is
unavoidable, since \textsc{AdaGrad MD} and similar algorithms based on
\textsc{MD} can have $\Omega(T)$ regret if Bregman divergence is not bounded.

@@

For unbounded decision sets no scale-free algorithm with a non-trivial regret
bound was known. Unbounded decision sets are practically important, since
learning large-scale linear models (e.g. logistic regression) is done by
reducing it to OLO with decision set $\R^d$.

@@

We prove that the regret of \textsc{AdaFTRL} and \textsc{SOLO FTRL} after $T$
rounds is bounded by $O \left(\sqrt{\sup_{v \in K} f(v)
\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $f$ is a non-negative regularizer
that is $1$-strongly convex with respect to a norm $\|\cdot\|$ and
$\|\cdot\|_*$ is its dual norm. For \textsc{Scale-Free MD}, we prove $O
\left(\sqrt{\sup_{u,v \in K} B_f(u,v) \sum_{t=1}^T\|\ell_t\|_*^2} \right)$
where $B_f$ is the Bregman divergence associated with a $1$-strongly convex
regularizer $f$.  As we discuss, in Section~\ref{section:mirror-descent}, the
upper bound for \textsc{Scale-Free MD} is never better than the bound for bound
for \textsc{AdaFTRL} and \textsc{SOLO FTRL}.  In
Section~\ref{section:lower-bound}, we show that the $\sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$ term in the bounds term is necessary by proving
$\frac{D}{\sqrt{8}} \sqrt{\sum_{t=1}^T\|\ell_t\|_*^2}$ lower bound on the
regret of any algorithm for OLO for any decision set with diameter $D$ with
respect to the primal norm $\|\cdot\|$.

For \textsc{SOLO FTRL}, we prove that regret against a competitor $u \in K$ is
at most $O \left(f(u) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} + \max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} \right)$.  Same as before, $f$ is a non-negative
$1$-strongly convex regularizer. This bound is non-trivial for any decision
set, bounded or unbounded.  The result makes \textsc{SOLO FTRL} the
\textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial regret bound.

@@

Our proof technique are also new because they rely on new homogeneous
inequalities (Lemmas~\ref{lemma:recurrence-solution}, \ref{lemma:useful}).

### 2015-12-13 limits.tex
Fix typos, grammar, and bibliography.

Before we give the proof, we illustrate the lemma on a simple
example.  Let $K = [-1,1]^2$ be a closed square in $\R^2$ and let
$R(w) = \|w\|_2^2$. Let $L = (1,0)$. The minimizers are
$$
\argmin_{w \in K} \langle L, w \rangle = \{ (-1,y) ~:~ y \in [-1,1] \} \; .
$$
The minimizer with smallest value of $R(w)$ is $(-1,0)$. Hence the lemma implies
that
$$
\lim_{\eta \to +\infty} \argmin_{w \in K}
\left( \langle L, w \rangle + \frac{1}{\eta} \|w\|_2^2 \right) = (-1, 0) \; .
$$

\begin{proof}
Without loss of generality, we can assume that $R(w)$ is non-negative for any
$w \in K$.  For otherwise, we can replace $R(w)$ with $R'(w) = R(w) - \inf_{v
\in K} R(v)$.

>>>>

Before we give the proof, we illustrate the lemma on a simple
example.  Let $K = [-1,1]^2$ be a closed square in $\R^2$ and let
$R(w) = \|w\|_2^2$. Let $L = (1,0)$. The minimizers are
$$
\argmin_{w \in K} \langle L, w \rangle = \{ (-1,y) ~:~ y \in [-1,1] \} \; .
$$
The minimizer with the smallest value of $R(w)$ is $(-1,0)$. Hence the lemma
implies that
$$
\lim_{\eta \to +\infty} \argmin_{w \in K}
\left( \langle L, w \rangle + \frac{1}{\eta} \|w\|_2^2 \right) = (-1, 0) \; .
$$

\begin{proof}[Proof of Lemma~\ref{lemma:prediction-limit-existence}]
Without loss of generality, we can assume that $R(w)$ is non-negative for any
$w \in K$.  For otherwise, we can replace $R(w)$ with $R'(w) = R(w) - \inf_{v
\in K} R(v)$.
